\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{natbib}
\usepackage[margin=1in]{geometry}

\DeclareMathOperator{\eml}{eml}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\bg}{\mathbf{g}}
\newcommand{\bK}{\mathbf{K}}
\newcommand{\bC}{\mathbf{C}}
\newcommand{\bV}{\mathbf{V}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bm}{\mathbf{m}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\cK}{\mathcal{K}}
\newcommand{\cM}{\mathcal{M}}
\newcommand{\cC}{\mathcal{C}}
\newcommand{\cL}{\mathcal{L}}

\title{Koopman Operator Lifting via EML Observable Dictionaries:\\
A PySINDy-Integrated Framework with CTF Benchmarking}
\author{Jim Beaver}
\date{April 2026 --- Working Draft}

\begin{document}
\maketitle

\begin{abstract}
We introduce a framework for learning interpretable Koopman operator
representations using the EML (Exp-Minus-Log) Sheffer operator
$\eml(x,y) = \exp(x) - \ln(y)$ as the basis for observable dictionaries.
Since every elementary function can be expressed as a binary tree of identical
EML nodes, parameterized EML trees provide a \emph{complete}, gradient-trainable
search space for Koopman eigenfunctions.  After Gumbel-softmax annealing and
weight snapping, the learned observables are exact symbolic expressions ---
yielding closed-form Koopman eigenfunctions that black-box methods cannot produce.
We replace transcendental function calls with Horner-form Taylor-series primitives,
reducing the entire EML operator to fused multiply-add instructions for maximum GPU
throughput.  The resulting model is packaged as a PySINDy-compatible feature library
and validated against the dynamicsai.org Common Task Framework (CTF) 12-metric
benchmark on the Lorenz system, comparing to EDMD, deep Koopman autoencoders, and
PySR baselines.
\end{abstract}

% ============================================================================
\section{Introduction}
\label{sec:intro}

The Koopman operator provides a linear (but infinite-dimensional) description of
nonlinear dynamical systems \citep{koopman1931,mezic2005,budisic2012}.  A central challenge is selecting
the dictionary of observable functions that span a finite-dimensional invariant
subspace.  Traditional approaches --- polynomials, radial basis functions, Fourier
modes, delay coordinates --- are chosen ad hoc and offer no completeness guarantee.
Deep-learning methods \citep{takeishi2017,lusch2018} learn flexible representations
but sacrifice interpretability: the resulting observables are opaque neural
networks with no closed-form expressions.  Symbolic regression
\citep{schmidt2009,cranmer2023} can discover closed-form expressions, but existing
approaches lack the completeness guarantee that EML provides.

\citet{odrzywołek2026} discovered that the binary operator
$\eml(x,y) = \exp(x) - \ln(y)$, paired with the constant~$1$, generates
\emph{every} standard elementary function --- the continuous-math analog of the NAND
gate \citep{sheffer1913} for Boolean logic.  This provides a natural, complete basis for Koopman
observable dictionaries.  Because every elementary function admits an EML tree
representation, the search space is guaranteed to contain the true Koopman
eigenfunctions whenever they are elementary --- which holds for a wide class of
physically-motivated systems (harmonic oscillators, Lotka--Volterra, pendula, etc.).

After training with Gumbel-softmax annealing \citep{jang2017}, the continuous routing parameters snap
to discrete one-hot selections, yielding \emph{exact symbolic expressions}.  The
resulting Koopman eigenfunctions are linear combinations of these snapped EML
observables and are therefore also elementary functions --- fully interpretable and
amenable to analytic manipulation.

\paragraph{Contributions.}
\begin{itemize}
    \item Parameterized EML trees as gradient-trainable Koopman observables with
          exact symbolic snap, providing a \emph{complete} dictionary over elementary
          functions.
    \item GPU-optimized Taylor-series EML primitives in Horner form with range
          reduction, achieving zero transcendental function calls in both forward
          and backward passes.
    \item A fully vectorized tree architecture that evaluates all dictionary elements
          in a single batched call per tree level.
    \item A three-phase training pipeline (exploration, hardening, snap \& fine-tune)
          with Gumbel-softmax annealing.
    \item PySINDy integration via \texttt{EMLLibrary}.
    \item CTF benchmark evaluation (E1--E12) on Lorenz, with comparison to EDMD,
          deep Koopman, and PySR baselines.
\end{itemize}


% ============================================================================
\section{Background}
\label{sec:background}

\subsection{The EML Operator}
\label{sec:eml}

\citet{odrzywołek2026} introduced a binary operator that, together with the
constant~$1$, generates every standard elementary function.  The operator is
defined as:
\begin{equation}\label{eq:eml}
  \eml(x, y) \;=\; \exp(x) - \ln(y).
\end{equation}
The grammar is context-free and trivially simple:
\begin{equation}\label{eq:grammar}
  S \;\to\; 1 \;\mid\; \eml(S, S).
\end{equation}
Every elementary expression therefore becomes a \emph{binary tree of identical
nodes}.  Key identities include:

\begin{center}
\begin{tabular}{lcl}
\toprule
Function & EML Form & Depth \\
\midrule
$e^x$ & $\eml(x, 1)$ & 1 \\
$e$ & $\eml(1, 1)$ & 1 \\
$\ln(x)$ & $\eml\!\bigl(1,\;\eml(\eml(1, x),\; 1)\bigr)$ & 3 \\
$-x$ & depth-5 composition & 5 \\
$x \times y$ & depth-8 composition & 8 \\
\bottomrule
\end{tabular}
\end{center}

Internally, computations proceed over~$\C$ via the principal branch of the complex
logarithm.  Trigonometric functions emerge through Euler's formula:
$e^{i\varphi} = \cos\varphi + i\sin\varphi$, with $i = \sqrt{-1}$ obtainable from
$\ln(-1) = i\pi$.


\subsection{Koopman Operator Theory}
\label{sec:koopman}

Consider a discrete-time nonlinear dynamical system on a state space
$\cM \subseteq \R^n$:
\begin{equation}\label{eq:dynamics}
  \bx_{k+1} = F(\bx_k).
\end{equation}
The \textbf{Koopman operator}~$\cK$ acts on scalar-valued observables
$g : \cM \to \C$ by composition with the dynamics:
\begin{equation}\label{eq:koopman}
  [\cK\, g](\bx) \;=\; g\bigl(F(\bx)\bigr).
\end{equation}
This is an infinite-dimensional linear operator, even when~$F$ is nonlinear.  The
key insight: if we find a finite set of observables
$\bg(\bx) = [g_1(\bx), \ldots, g_N(\bx)]^\top$ that span a $\cK$-invariant
subspace, then the dynamics in lifted coordinates become:
\begin{equation}\label{eq:lifted}
  \bg(\bx_{k+1}) \;=\; \bK \;\bg(\bx_k),
\end{equation}
where $\bK \in \C^{N \times N}$ is a finite-dimensional \textbf{Koopman matrix}.
The central problem is choosing the dictionary of observables~$\{g_j\}$.

\textbf{Extended Dynamic Mode Decomposition (EDMD)} \citep{williams2015} solves
for~$\bK$ by least-squares regression given a fixed, hand-selected dictionary.
The original DMD algorithm \citep{schmid2010,tu2014} uses linear observables;
EDMD extends this to nonlinear dictionaries.  Typical choices --- polynomials,
radial basis functions, Fourier modes --- are ad hoc and offer no guarantee that
the relevant invariant subspace is spanned.  Kernel EDMD
\citep{williams2015kernel} and deep Koopman autoencoders \citep{lusch2018,otto2019}
address expressiveness but sacrifice either tractability or interpretability.
For a comprehensive review see \citet{brunton2022modern}.


\subsection{SINDy and PySINDy}
\label{sec:sindy}

The Sparse Identification of Nonlinear Dynamics (SINDy)
framework \citep{brunton2016} discovers governing equations by regressing
time derivatives against a library of candidate nonlinear functions using
sparsity-promoting optimization.  PySINDy \citep{desilva2020,kaptanoglu2022}
provides a modular Python implementation with interchangeable feature libraries.

Our approach connects to SINDy through the shared notion of a feature library:
rather than using pre-defined polynomial or Fourier features, we supply EML trees
as the library.  After symbolic snap, the discovered features can be fed into a
standard SINDy regression pipeline, combining EML's completeness guarantee with
SINDy's sparse model selection.


% ============================================================================
\section{Method}
\label{sec:method}

\subsection{EML Trees as Koopman Observables}
\label{sec:eml-trees}

\subsubsection{The Master Formula}

Each input to an EML node is parameterized as a soft selection among a candidate
set:
\begin{equation}\label{eq:master}
  \text{input}_i \;=\; \alpha_i + \beta_i \cdot x + \gamma_i \cdot f_{\text{child}},
\end{equation}
where $(\alpha_i, \beta_i, \gamma_i)$ are trainable weights that select among:

\begin{center}
\begin{tabular}{lcccc}
\toprule
Selection & $\alpha_i$ & $\beta_i$ & $\gamma_i$ & Result \\
\midrule
Constant $1$ & 1 & 0 & 0 & $1$ \\
Input variable & 0 & 1 & 0 & $x$ \\
Child output & 0 & 0 & 1 & $f_{\text{child}}$ \\
\bottomrule
\end{tabular}
\end{center}

During training, $(\alpha, \beta, \gamma)$ are treated as \textbf{Gumbel-softmax
logits} \citep{jang2017} normalized to a simplex.  Temperature~$\tau$ anneals from soft
(exploration) to hard (discrete).  After convergence, the weights snap to one-hot
$\{0, 1\}$, yielding an exact symbolic expression.

More generally, at each EML node~$i$, both left and right inputs are selected from
a candidate set
$\cC_i = \{1, x_0, x_1, \ldots, x_{n-1}\}$ (for leaves) or
$\cC_i = \{1, x_0, \ldots, x_{n-1}, f_{\text{child}}\}$ (for interior nodes).
The selection is parameterized by logits $\ell_i \in \R^{|\cC_i|}$:
\begin{equation}\label{eq:gumbel}
  w_i^{(j)} = \frac{\exp\!\bigl((\ell_i^{(j)} + G_j)/\tau\bigr)}
  {\sum_k \exp\!\bigl((\ell_i^{(k)} + G_k)/\tau\bigr)},
  \qquad G_j \sim \text{Gumbel}(0,1).
\end{equation}
The selected input is $\sum_j w_i^{(j)} c_j$ where $c_j \in \cC_i$.  As
$\tau \to 0$, this converges to hard one-hot selection.

\subsubsection{Full Level-2 Example}

The complete level-2 master formula (14 parameters for univariate~$x$) is:
\begin{equation}\label{eq:level2}
\begin{split}
  F(x) = \eml\Bigl[\;
    &\alpha_1 + \beta_1 x + \gamma_1\,\eml(\alpha_3 + \beta_3 x,\;\alpha_4 + \beta_4 x), \\
    &\alpha_2 + \beta_2 x + \gamma_2\,\eml(\alpha_5 + \beta_5 x,\;\alpha_6 + \beta_6 x)
  \;\Bigr].
\end{split}
\end{equation}
Setting $\alpha_1\!=\!0, \beta_1\!=\!1, \gamma_1\!=\!0$ and
$\alpha_2\!=\!1, \beta_2\!=\!\gamma_2\!=\!0$ recovers $\exp(x)$.  The general
depth-$d$ formula has $5 \cdot 2^d - 6$ parameters.

\subsubsection{Koopman Dictionary Construction}

We construct $N$ independent EML trees, each of depth~$d$, to form the observable
dictionary:
\begin{equation}\label{eq:dictionary}
  \bg_\theta(\bx) \;=\; \bigl[\,T_1^\theta(\bx),\;\; T_2^\theta(\bx),\;\;
  \ldots,\;\; T_N^\theta(\bx)\,\bigr]^\top,
\end{equation}
where $T_j^\theta$ denotes the $j$-th EML tree with trainable routing
parameters~$\theta$.  The full Koopman model is then:
\begin{equation}\label{eq:koopman-model}
  \bg_\theta(\bx_{k+1}) \;\approx\; \bK\;\bg_\theta(\bx_k), \qquad
  \bx_k \;\approx\; \bC\;\bg_\theta(\bx_k),
\end{equation}
where $\bK \in \R^{N \times N}$ is the Koopman matrix and
$\bC \in \R^{n \times N}$ is the decoder (reconstruction map).

\subsubsection{Vectorized Tree Architecture}

All $N$ trees are packed into contiguous parameter tensors, one per tree level:
\begin{equation}
  \text{level\_logits}[l] \;\in\; \R^{N \times n_l \times |\cC_l| \times 2},
\end{equation}
where $n_l = 2^{d-1-l}$ is the number of nodes at level~$l$ and the trailing
dimension~2 indexes left/right inputs.  Bottom-up evaluation proceeds as $d$
sequential batched operations (Algorithm~\ref{alg:tree-eval}).

\begin{algorithm}[t]
\caption{Vectorized EML Tree Evaluation}
\label{alg:tree-eval}
\begin{algorithmic}[1]
\REQUIRE State batch $\bx \in \R^{B \times n}$, tree depth $d$, temperature $\tau$
\ENSURE Lifted observables $\bg \in \R^{B \times N}$
\FOR{$l = 0, 1, \ldots, d-1$ (bottom to top)}
  \STATE Compute Gumbel-softmax weights
    $W \in \R^{B \times N \times n_l \times |\cC_l| \times 2}$
  \STATE Assemble candidate tensors (const $1$, input vars, child outputs from
    level $l\!-\!1$)
  \STATE Weighted selection:
    $\text{eml\_x}, \text{eml\_y} \in \R^{B \times N \times n_l}$
  \STATE Clamp inputs:
    $\text{eml\_x} \gets \text{clamp}(\text{eml\_x}, -C, C)$
  \STATE Batched EML:
    $\text{out}_l = \eml_{\text{Taylor}}(\text{eml\_x},\; \text{eml\_y})$
\ENDFOR
\RETURN $\text{out}_{d-1} \in \R^{B \times N}$ (root outputs = lifted state)
\end{algorithmic}
\end{algorithm}

\subsubsection{Parameter Counts}

The model is compact by modern standards:

\begin{center}
\begin{tabular}{ccccc}
\toprule
Depth $d$ & Params/tree & $N\!=\!16$ & $N\!=\!32$ & Expressible functions \\
\midrule
2 & 14 & 224 & 448 & $e^x$, $\ln x$, $e^{e^x}$ \\
3 & 34 & 544 & 1,088 & $-x$, $1/x$, most unary \\
4 & 74 & 1,184 & 2,368 & $x+y$, $x \cdot y$, trig \\
5 & 154 & 2,464 & 4,928 & Full calculator \\
\bottomrule
\end{tabular}
\end{center}

\noindent
\citet{odrzywołek2026} reports ${\sim}25\%$ blind recovery at depth 3--4,
improving to $100\%$ with warm-start initialization.


\subsection{Taylor-Series GPU Primitives}
\label{sec:taylor}

Both $\exp$ and $\ln$ can be replaced by polynomial approximations in Horner form,
reducing the entire EML operator to fused multiply-add (FMA) instructions with zero
transcendental function calls.  This is critical for GPU throughput, especially with
nested tree compositions where each node calls EML.

\subsubsection{Taylor Exponential with Range Reduction}

Decompose $x = k \ln 2 + r$ where $k = \lfloor x / \ln 2 \rceil$ and
$|r| \le \frac{\ln 2}{2} \approx 0.347$.  Then:
\begin{equation}\label{eq:exp-range}
  \exp(x) = 2^k \cdot \exp(r), \qquad
  \exp(r) = \sum_{n=0}^{M} \frac{r^n}{n!}.
\end{equation}
The Taylor series is evaluated in Horner form (innermost term outward):
\begin{equation}\label{eq:exp-horner}
  \exp(r) \;=\; 1 + r\!\left(1 + \frac{r}{2}\!\left(1 + \frac{r}{3}\!\left(1
  + \cdots \frac{r}{M}\right)\!\right)\!\right).
\end{equation}
Each step is one multiply-add.  With $|r| \le 0.347$, order $M = 12$ gives error
$\lesssim 10^{-16}$ (double precision).  The $2^k$ scaling is a single
\texttt{ldexp}-equivalent operation.

\subsubsection{Taylor Logarithm via Arctanh}

Range-reduce: $y = m \cdot 2^e$ with $m \in [0.5, 1.0)$, so
$\ln(y) = e \ln 2 + \ln(m)$.  Then use the arctanh series, which converges much
faster than the naive $\ln(1+t)$ expansion:
\begin{equation}\label{eq:ln-arctanh}
  \ln(m) \;=\; 2\,\operatorname{arctanh}\!\left(\frac{m-1}{m+1}\right)
  \;=\; 2\sum_{k=0}^{M}\frac{u^{2k+1}}{2k+1}, \quad u = \frac{m-1}{m+1}.
\end{equation}
For $m \in [0.5, 1)$ we have $|u| \le \tfrac{1}{3}$, giving rapid convergence.
Horner evaluation:
\begin{equation}\label{eq:ln-horner}
  \sum_{k=0}^{M} \frac{u^{2k+1}}{2k+1} \;=\; u\!\left(\frac{1}{1} +
  u^2\!\left(\frac{1}{3} + u^2\!\left(\frac{1}{5} + \cdots\right)\!\right)\!\right).
\end{equation}
Order $M = 16$ gives machine-epsilon accuracy for $\ln$.

\subsubsection{Combined EML Primitive}

The GPU-optimized EML primitive is:
\begin{equation}\label{eq:eml-taylor}
  \eml_{\text{Taylor}}(x, y) \;=\; \text{horner\_exp}(x, M_e) \;-\;
  \text{horner\_ln}(y, M_l),
\end{equation}
with gradients $\frac{\partial}{\partial x}\eml = \exp(x)$ and
$\frac{\partial}{\partial y}\eml = -\frac{1}{y}$, both computable via the same
polynomial machinery.  A custom \texttt{torch.autograd.Function} \citep{paszke2019} avoids redundant
recomputation in the backward pass: the forward stores $x$ and $y$ and reuses the
Horner evaluation for gradient computation.

For a depth-$d$ tree with $N$ parallel observables evaluated over a batch of
size~$B$, the forward pass requires $d$ sequential EML calls, each operating on
tensors of shape $[B, N, 2^{d-1-l}]$ at level~$l$.  All nodes at the same depth
are computed in a single batched call --- no Python loops over individual nodes.


\subsection{PySINDy Integration}
\label{sec:pysindy}

We package the EML dictionary as a PySINDy-compatible feature library,
\texttt{EMLLibrary}, supporting two modes:

\begin{itemize}
  \item \textbf{Enumerative mode:} Generate all distinct EML trees up to a specified
    depth (filtered by algebraic equivalence).  This provides a finite but complete
    basis for SINDy regression at a chosen complexity level.
  \item \textbf{Pre-trained mode:} Supply trees already trained via
    Gumbel-softmax Koopman optimization (Section~\ref{sec:training}).
    After snapping, each tree is a fixed symbolic feature.  SINDy's
    sparsity-promoting regression then selects the most relevant subset.
\end{itemize}

\noindent
Both modes implement the standard PySINDy \texttt{BaseFeatureLibrary} interface,
allowing seamless composition with existing PySINDy optimizers (STLSQ, SR3, etc.).


% ============================================================================
\section{Experimental Setup}
\label{sec:experiments}

\subsection{Training Pipeline}
\label{sec:training}

\subsubsection{Loss Function}

Given trajectory data $\{(\bx_k, \bx_{k+1})\}_{k=1}^{K}$, the joint objective is:
\begin{equation}\label{eq:loss}
  \cL(\theta, \bK, \bC) \;=\;
  \underbrace{\sum_k \bigl\|\bg_\theta(\bx_{k+1}) - \bK\,\bg_\theta(\bx_k)\bigr\|^2}%
  _{\text{prediction}}
  \;+\; \lambda\,
  \underbrace{\sum_k \bigl\|\bx_k - \bC\,\bg_\theta(\bx_k)\bigr\|^2}%
  _{\text{reconstruction}}
  \;+\; \mu\,
  \underbrace{\|\bK\|_F^2}_{\text{regularization}}.
\end{equation}
All three components are differentiable through the EML trees (the Taylor
polynomials have clean analytic gradients).

\subsubsection{Three-Phase Annealing Schedule}

Training proceeds in three phases:

\begin{enumerate}
  \item \textbf{Phase 1 --- Exploration} (epochs $0$ to $0.6T$):
    Anneal $\tau$ from $\tau_{\max}$ (e.g., 2.0) to~$1.0$.  All parameters
    $(\theta, \bK, \bC)$ are trainable.  Gumbel noise encourages exploration of
    different routing configurations.
  \item \textbf{Phase 2 --- Hardening} (epochs $0.6T$ to $0.9T$):
    Anneal $\tau$ from $1.0$ to $\tau_{\min}$ (e.g., 0.1).  Routing weights sharpen
    toward discrete selections; the trees commit to specific functional forms.
  \item \textbf{Phase 3 --- Snap \& Fine-Tune} (epochs $0.9T$ to $T$):
    Snap all routing logits to hard one-hot (argmax).  Freeze tree
    parameters~$\theta$.  Fine-tune only $\bK$ and $\bC$ at reduced learning rate.
\end{enumerate}

\noindent
\textbf{Overflow hazard.}  Nested EML compositions produce iterated exponentials
($e^{e^{e^x}}$) during exploration.  EML inputs are clamped to $[-C, C]$ with
$C = 5$ (so $\exp(C) \approx 148$), and state data is normalized to zero-mean
unit-variance before lifting.  The tight clamp is essential: $C = 20$ yields
$\exp(20) \approx 5 \times 10^8$, causing catastrophic loss magnitudes in early
training.

\subsubsection{Warm-Start Strategies}

Convergence from random initialization degrades sharply beyond depth~4.  Two
warm-start strategies improve recovery:

\begin{enumerate}
  \item \textbf{EDMD bootstrap:} Run standard EDMD with a polynomial dictionary
    first.  Initialize EML tree routing to mimic the best polynomial observables
    (since polynomials are elementary, they have EML representations).
  \item \textbf{Curriculum depth:} Train at depth~2 until convergence, then grow
    each tree by one level (appending new leaf nodes with random logits) and
    continue training.  The existing structure provides a strong initialization for
    the expanded tree.
\end{enumerate}


\subsection{CTF Benchmark Framework}
\label{sec:ctf}

We evaluate on the dynamicsai.org Common Task Framework
(CTF) \citep{wyder2025}, which defines
12 standardized metrics (E1--E12) for dynamical-system learning.  Each metric
follows the scoring formula
$E = 100 \times (1 - \|\text{pred} - \text{truth}\| / \|\text{truth}\|)$,
where $E = 100$ is perfect and $E < 0$ is worse than trivial.
For long-term metrics the comparison is made in log-power-spectrum space.
The 12 metrics are grouped into four categories:

\begin{itemize}
  \item \textbf{Forecasting (E1--E2):}
    E1 evaluates short-term trajectory tracking (normalized RMSE over
    1--5 Lyapunov times).
    E2 evaluates long-term spectral fidelity (log power spectrum error).

  \item \textbf{Noisy data (E3--E6):}
    E3/E4 test reconstruction and forecasting from medium-noise observations
    (SNR~$\approx 20$\,dB).
    E5/E6 repeat at high noise (SNR~$\approx 10$\,dB).

  \item \textbf{Limited data (E7--E10):}
    E7/E8 evaluate short- and long-term performance when trained on only
    10\% of the clean data.
    E9/E10 repeat with 10\% noisy data.

  \item \textbf{Parametric generalization (E11--E12):}
    E11 tests interpolation: train at parameter values bracketing the test
    value (e.g., $\rho \in \{26, 30\}$, test at $\rho = 28$).
    E12 tests extrapolation: train at $\rho \in \{26, 28\}$, test at
    $\rho = 32$.
\end{itemize}

\noindent
We use the Lorenz-63 system ($\sigma\!=\!10$, $\rho\!=\!28$, $\beta\!=\!8/3$) as
the primary test case, with 20,000 training pairs sampled at $\Delta t = 0.01$
from 20 trajectories normalized to zero-mean unit-variance.


\subsection{Baselines}
\label{sec:baselines}

We compare four methods:

\begin{enumerate}
  \item \textbf{EML-Koopman (ours):} $N\!=\!16$ trees at depth~2 (720
    parameters), three-phase training with $T\!=\!1200$ epochs, batch
    size~2048, input clamping at $\pm 5$.  Taylor orders: $M_e\!=\!10$,
    $M_l\!=\!12$.
  \item \textbf{EDMD-Poly:} Polynomial dictionary up to degree~4
    ($\binom{n+4}{4} = 35$ features for $n\!=\!3$), least-squares $\bK$.
  \item \textbf{EDMD-RBF:} 200 Gaussian RBF centers selected randomly from the
    training data, $\sigma\!=\!1.0$.
  \item \textbf{Deep Koopman:} 3-layer MLP autoencoder ($3 \to 128 \to 128
    \to 32$), symmetric decoder, linear dynamics in 32-dimensional latent
    space (43{,}299 parameters), trained for 1200 epochs.
\end{enumerate}


% ============================================================================
\section{Results}
\label{sec:results}

\subsection{CTF Lorenz Benchmark}

Table~\ref{tab:ctf} summarizes the CTF Lorenz benchmark results for all methods.
All models were trained on 20,000 state pairs from 20 Lorenz trajectories
($\sigma\!=\!10$, $\rho\!=\!28$, $\beta\!=\!8/3$, $\Delta t\!=\!0.01$),
normalized to zero-mean unit-variance.  Forecasting is evaluated on a held-out
trajectory segment (501 time steps).

\begin{table}[t]
\centering
\caption{CTF Lorenz benchmark comparison.  E1 = short-term forecasting score
  ($100 \times (1 - \text{NRMSE})$, higher is better).  E2 = long-term spectral
  score (higher is better).  Valid steps = number of time steps before prediction
  error exceeds $0.5\sigma$ of the test trajectory.  Published CTF leaderboard
  values (E1--E12 average) shown for context.}
\label{tab:ctf}
\begin{tabular}{lccccc}
\toprule
Method & E1 & E2 & RMSE & Valid Steps & Params \\
\midrule
\textbf{EML-Koopman (ours)} & \textbf{7.70} & $-12.23$ & \textbf{0.888} & \textbf{146} & 720 \\
EDMD-Poly (degree 4) & $-1.11$ & \textbf{52.94} & 0.973 & 84 & $\sim$35 \\
EDMD-RBF (200 centers) & $-10{,}216$ & $-252$ & 99.3 & 0 & $\sim$200 \\
Deep Koopman (autoencoder) & 0.63 & $-34.28$ & 0.956 & 105 & 43{,}299 \\
\midrule
LSTM (CTF published) & \multicolumn{2}{c}{avg = 64.54} & --- & --- & --- \\
DeepONet (CTF published) & \multicolumn{2}{c}{avg = 57.80} & --- & --- & --- \\
Reservoir (CTF published) & \multicolumn{2}{c}{avg = 54.87} & --- & --- & --- \\
KAN (CTF published) & \multicolumn{2}{c}{avg = 47.28} & --- & --- & --- \\
SINDy (CTF published) & \multicolumn{2}{c}{avg = $-3.00$} & --- & --- & --- \\
PyKoopman (CTF published) & \multicolumn{2}{c}{avg = $-20.11$} & --- & --- & --- \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Short-term forecasting (E1).}
EML-Koopman achieves E1 $= 7.70$, the highest among all implemented baselines and
substantially above the two other interpretable methods on the CTF leaderboard:
SINDy ($-3.00$) and PyKoopman ($-20.11$).  The valid prediction horizon of 146
time steps (1.46 Lyapunov times at $\Delta t = 0.01$) exceeds EDMD-Poly (84 steps)
and Deep Koopman (105 steps) despite using 60$\times$ fewer parameters than the
autoencoder.  The gap to black-box methods (LSTM 64.54, DeepONet 57.80) reflects
the expected trade-off between parameter capacity and interpretability.

\paragraph{Long-term statistics (E2).}
EDMD-Poly achieves the highest E2 ($52.94$) by a wide margin, because its
polynomial basis functions naturally capture the power spectral shape of the
Lorenz attractor even when short-term prediction diverges.  EML-Koopman's E2
of $-12.23$ indicates room for improvement in long-term spectral fidelity.
A systematic depth sweep (Section~\ref{sec:depth-analysis}) investigates whether
deeper EML trees can close this gap.

\paragraph{Failure modes.}
EDMD-RBF diverges catastrophically (E1 $= -10{,}216$) because random Gaussian
centers in the normalized 3D Lorenz state space produce a severely ill-conditioned
Koopman matrix.  This underscores the importance of dictionary design --- a problem
that EML addresses through gradient-based observable discovery.

\subsection{Discovered Symbolic Observables}

After Phase~3 snap, all 16 EML tree observables reduce to closed-form
elementary functions.  Representative examples (with $x_0, x_1, x_2$
corresponding to normalized Lorenz $x, y, z$):

\begin{center}
\begin{tabular}{cl}
\toprule
Observable & Snapped Formula \\
\midrule
$g_0$ & $\eml(\eml(1, x_1),\; x_1)$ \\
$g_1$ & $\eml(\eml(1, x_1),\; 1)$ \\
$g_2$ & $\eml(\eml(1, x_0),\; 1)$ \\
$g_4$ & $\eml(\eml(1, x_1),\; x_2)$ \\
$g_{10}$ & $\eml(\eml(1, x_0),\; x_1)$ \\
\bottomrule
\end{tabular}
\end{center}

\noindent
These simplify to:
\begin{align*}
  g_1(\bx) &= \exp(e - \ln x_1) = e \cdot x_1^{-1} \cdot \exp(0) \quad
    \text{(involves $1/y$)}, \\
  g_2(\bx) &= \exp(e - \ln x_0) \quad
    \text{(involves $1/x$)}, \\
  g_4(\bx) &= \exp(e - \ln x_1) - \ln x_2 \quad
    \text{(mixes exponential and logarithmic)}.
\end{align*}
The model predominantly discovers observables built from $\eml(1, x_i) = e - \ln x_i$
as an inner expression, then applies a second EML layer.  This structure is natural:
depth-2 trees can represent $e^x$, $\ln x$, and their compositions, but not
yet multiplication or addition (which require depth $\ge 5$).  The abundance of
logarithmic and exponential-reciprocal features suggests the model has found the
most informative nonlinear transformations available at depth~2.

\paragraph{Interpretability advantage.}
Unlike the 43{,}299-parameter Deep Koopman autoencoder --- whose observables are
opaque 3-layer neural networks --- every EML observable is a human-readable formula.
The resulting Koopman eigenfunctions $\varphi_j(\bx) = \bv_j^\top \bg(\bx)$ are
therefore explicit linear combinations of elementary functions, amenable to analytic
simplification and physical interpretation.


\subsection{Progressive Depth Analysis}
\label{sec:depth-analysis}

A key hypothesis motivating this work is that deeper EML trees --- which can express
a richer class of elementary functions --- should improve Koopman approximation
quality.  To test this, we swept tree depths $d \in \{2, 3, 4, 5\}$ with multiple
observable counts $N \in \{16, 24, 32\}$ per depth, training each configuration
from random initialization with depth-adapted hyperparameters (tighter input
clamping and lower learning rates at greater depths to manage numerical instability
from iterated exponentials).  Table~\ref{tab:depth} summarizes all 11
configurations.

\begin{table}[t]
\centering
\caption{Progressive depth sweep on CTF Lorenz.  Each row is a separate training
  run from random initialization.  Clamp = symmetric input clamp $[-C, C]$ applied
  before each EML call.  Best E1 and best E2 among EML configurations are bolded.}
\label{tab:depth}
\begin{tabular}{lccccccccc}
\toprule
Config & $d$ & $N$ & Clamp & Epochs & E1 & E2 & RMSE & \shortstack{Valid\\Steps} & Params \\
\midrule
\textbf{d2-N16} & 2 & 16 & 5.0 & 1200 & \textbf{7.70} & \textbf{$-$12.23} & \textbf{0.888} & \textbf{146} & 720 \\
d2-N24 & 2 & 24 & 5.0 & 1500 & 7.84 & $-$47.40 & 0.887 & 146 & 1{,}272 \\
d2-N32 & 2 & 32 & 5.0 & 1500 & 7.20 & $-$70.24 & 0.893 & 144 & 1{,}952 \\
\midrule
d3-N16 & 3 & 16 & 4.0 & 1500 & 5.32 & $-$48.58 & 0.911 & 143 & 1{,}296 \\
d3-N24 & 3 & 24 & 4.0 & 2000 & 4.08 & $-$37.29 & 0.923 & 140 & 2{,}136 \\
d3-N32 & 3 & 32 & 4.0 & 2000 & 1.86 & $-$39.15 & 0.944 & 134 & 3{,}104 \\
\midrule
d4-N16 & 4 & 16 & 3.0 & 2000 & 2.09 & $-$68.88 & 0.942 & 134 & 2{,}448 \\
d4-N24 & 4 & 24 & 3.0 & 2500 & 4.56 & $-$24.67 & 0.918 & 141 & 3{,}864 \\
d4-N32 & 4 & 32 & 3.0 & 2500 & 3.43 & $-$27.71 & 0.929 & 140 & 5{,}408 \\
\midrule
d5-N16 & 5 & 16 & 2.5 & 2500 & 2.35 & $-$14.22 & 0.940 & 136 & 4{,}752 \\
d5-N24 & 5 & 24 & 2.5 & 3000 & 4.53 & $-$22.82 & 0.919 & 144 & 7{,}320 \\
\midrule
\multicolumn{5}{l}{EDMD-Poly (reference)} & $-$1.11 & 52.94 & 0.973 & 84 & $\sim$35 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Deeper trees do not improve E2.}
Contrary to the initial hypothesis, the best E2 ($-12.23$) comes from the
\emph{shallowest} configuration (d2-N16, 720 parameters).  Deeper trees
consistently produce worse long-term spectral fidelity, with E2 ranging from
$-14.22$ (d5-N16) to $-70.24$ (d2-N32).  Similarly, short-term E1 degrades
from 7.70 at depth~2 to as low as 1.86 at depth~3, indicating that the
optimization landscape becomes significantly harder at greater depths.

\paragraph{Formula collapse at depth~3+.}
Inspection of the snapped formulas reveals a structural explanation.
Table~\ref{tab:depth-formulas} shows representative observables from each depth.
At depth~2, the optimizer discovers diverse two-layer compositions such as
$\eml(\eml(1, x_1), x_0)$ and $\eml(\eml(1, x_0), x_2)$ that mix multiple
state variables through exponential and logarithmic transformations.  At depth~3
and beyond, most trees collapse to shallow forms --- typically
$\eml(1, x_i) = e - \ln x_i$, using only the root node while the deeper
structure goes unused.  This collapse suggests that gradient signal attenuates
through nested clamped exponentials, preventing the optimizer from coordinating
routing decisions across multiple tree levels.

\begin{table}[t]
\centering
\caption{Representative snapped EML formulas at each tree depth, illustrating
  the progressive collapse of deeper trees toward simple root-only expressions.
  Variables $x_0, x_1, x_2$ correspond to normalized Lorenz $x, y, z$.}
\label{tab:depth-formulas}
\begin{tabular}{clp{7.2cm}}
\toprule
Depth & Observable & Snapped Formula \\
\midrule
\multirow{4}{*}{2}
  & $g_0$ & $\eml\bigl(\eml(1, x_1),\; \eml(1, x_1)\bigr)$ \\
  & $g_1$ & $\eml\bigl(\eml(1, x_1),\; x_1\bigr)$ \\
  & $g_3$ & $\eml\bigl(\eml(1, x_1),\; x_0\bigr)$ \\
  & $g_6$ & $\eml\bigl(\eml(1, x_0),\; x_2\bigr)$ \\
\midrule
\multirow{3}{*}{3}
  & $g_0$ & $\eml(1,\; x_0)$ \\
  & $g_1$ & $\eml(1,\; x_1)$ \\
  & $g_6$ & $\eml(1,\; x_2)$ \\
\midrule
\multirow{3}{*}{4}
  & $g_0$ & $\eml(1,\; x_0)$ \\
  & $g_3$ & $\eml(x_1,\; x_1)$ \\
  & $g_6$ & $\eml(x_1,\; x_0)$ \\
\midrule
\multirow{3}{*}{5}
  & $g_0$ & $\eml(1,\; x_1)$ \\
  & $g_1$ & $\eml(x_2,\; x_0)$ \\
  & $g_2$ & $\eml(x_1,\; x_0)$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Structural barrier to E2.}
The persistent 65-point E2 gap between EML-Koopman ($-12.23$) and EDMD-Poly
($52.94$) has a structural explanation.  EDMD-Poly's 4th-degree polynomial
dictionary directly contains monomials ($x^2$, $xy$, $x^2 z$, etc.) that
closely approximate the Lorenz attractor's power spectral shape.  In EML tree
representation, \emph{addition} requires depth~$\ge 5$ and \emph{multiplication}
requires depth~$\ge 8$ \citep{odrzywołek2026}.  The depth-2 trees used in our
best configuration cannot express any polynomial of degree $\ge 2$; their
observable space is restricted to compositions of $\exp$ and $\ln$ --- functions
whose spectral signatures differ fundamentally from the polynomial features that
capture Lorenz long-term statistics.  Closing this gap requires either much
deeper trees (with correspondingly harder optimization) or a hybrid dictionary
combining EML and polynomial features.

\paragraph{Fewer observables yield better E2.}
Within each depth, the smallest dictionary ($N\!=\!16$) consistently achieves
the best or near-best E2.  This suggests that over-parameterized dictionaries
introduce redundant or near-degenerate observables whose spectral contributions
interfere destructively in the long-term regime.


\subsection{Full CTF E1--E12 Evaluation}
\label{sec:full-ctf}

We evaluate EML-Koopman (d2-N16, 720 parameters) across all 12 CTF metrics
by constructing the required data scenarios: noisy observations at 20\,dB and
10\,dB SNR (E3--E6), 10\% training data subsets (E7--E10), and parametric
variations of the Lorenz $\rho$ parameter (E11--E12).  For noisy-data metrics,
the model is trained on clean data and the Koopman lift$\to$reconstruct pipeline
acts as a learned denoiser; for limited-data and parametric metrics, separate
models are trained on the reduced or multi-parameter datasets.
Table~\ref{tab:full-ctf} reports all 12 scores.

\begin{table}[t]
\centering
\caption{Full CTF E1--E12 scorecard for EML-Koopman (depth~2, $N\!=\!16$, 720
  parameters) on the Lorenz-63 system.  Scores follow the CTF formula:
  $E = 100 \times (1 - \|\text{pred} - \text{truth}\|/\|\text{truth}\|)$.
  Long-term metrics (E2, E4, E6, E8, E10) are computed in log-power-spectrum
  space.  Published CTF leaderboard averages are shown for comparison.}
\label{tab:full-ctf}
\begin{tabular}{clr}
\toprule
Metric & Category & Score \\
\midrule
E1  & Short-term forecasting            &   6.38 \\
E2  & Long-term spectral                &   3.77 \\
\midrule
E3  & Medium-noise reconstruction       &  33.68 \\
E4  & Medium-noise long-term forecast   & $-$41.87 \\
E5  & High-noise reconstruction         &  30.65 \\
E6  & High-noise long-term forecast     & $-$41.84 \\
\midrule
E7  & Limited data, short-term          &   7.77 \\
E8  & Limited data, long-term           & $-$47.37 \\
E9  & Limited noisy data, short-term    &   3.43 \\
E10 & Limited noisy data, long-term     &  $-$0.18 \\
\midrule
E11 & Parametric interpolation          &   1.16 \\
E12 & Parametric extrapolation          &  $-$0.12 \\
\midrule
    & \textbf{Average (E1--E12)}        & $\mathbf{-3.71}$ \\
\bottomrule
\end{tabular}

\vspace{4pt}
\begin{tabular}{lr}
\toprule
CTF Published (avg) & Avg Score \\
\midrule
LSTM        &  64.54 \\
DeepONet    &  57.80 \\
Reservoir   &  54.87 \\
KAN         &  47.28 \\
ODE-LSTM    &  41.67 \\
SINDy       & $-$3.00 \\
\textbf{EML-Koopman (ours)} & $\mathbf{-3.71}$ \\
PyKoopman   & $-$20.11 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Denoising via Koopman lifting (E3, E5).}
The most striking result is the strong performance on noisy-data reconstruction:
E3 $= 33.68$ at 20\,dB SNR and E5 $= 30.65$ at 10\,dB SNR.  The EML observable
dictionary spans a low-dimensional manifold (16 observables in 3D state space);
lifting noisy observations into this manifold and decoding back to state space
naturally rejects noise that lies outside the learned subspace.  This is
analogous to denoising via projection onto a low-rank subspace, but the
nonlinear EML lifting provides a richer manifold geometry than linear PCA.
The modest degradation from 20\,dB to 10\,dB (3 points) indicates robust
denoising even under severe noise contamination.

\paragraph{Robustness under data scarcity (E7).}
Training on only 10\% of the clean data yields E7 $= 7.77$, which
\emph{exceeds} the full-data E1 $= 6.38$.  This counter-intuitive result
reflects the strong inductive bias of the 720-parameter model: with fewer
training pairs, the model is naturally regularized and avoids overfitting to
the training trajectories.  The E8 long-term score ($-47.37$) remains negative,
consistent with the structural spectral limitation discussed above.

\paragraph{Noisy limited-data regime (E9--E10).}
Under simultaneous noise and data scarcity, EML-Koopman still produces positive
short-term scores (E9 $= 3.43$) and avoids catastrophic failure in the
long-term metric (E10 $= -0.18$, near zero rather than deeply negative).
This graceful degradation is attributable to the compact model size: with only
720 parameters, there is insufficient capacity to overfit noisy training data.

\paragraph{Parametric generalization (E11--E12).}
Interpolation (E11 $= 1.16$) is marginally positive and extrapolation
(E12 $= -0.12$) is near zero.  The learned Koopman operator~$\bK$ encodes
dynamics at specific parameter values and does not transfer across the
bifurcation parameter~$\rho$ without explicit parameterization.  Addressing
this limitation would require parameter-conditioned architectures, e.g.,
making the Koopman matrix a function of~$\rho$: $\bK(\rho)$.

\paragraph{Long-term spectral metrics.}
E4, E6, and E8 are all significantly negative ($-41$ to $-47$), consistent
with the E2 analysis in Section~\ref{sec:depth-analysis}: the exp/ln
observables at depth~2 cannot reproduce the Lorenz power spectrum shape
that polynomial bases capture naturally.

\paragraph{Overall positioning.}
The E1--E12 average of $-3.71$ places EML-Koopman within one point of
SINDy ($-3.00$) and 16 points above PyKoopman ($-20.11$) on the CTF
leaderboard --- competitive with the best interpretable methods while
providing closed-form symbolic observables.  The gap to black-box
methods (LSTM at 64.54, DeepONet at 57.80) reflects the fundamental
trade-off between the 720-parameter symbolic model and neural networks
with 10K--100K+ parameters.  Notably, EML-Koopman's denoising scores
(E3, E5) represent a distinctive capability not typically associated
with Koopman methods, suggesting that the nonlinear manifold structure
of EML dictionaries has practical value beyond forecasting accuracy.


% ============================================================================
\section{Discussion}
\label{sec:discussion}

\subsection{Eigendecomposition and Interpretability}

After training and snapping, the explicit Koopman matrix~$\bK$ admits
eigendecomposition:
\begin{equation}\label{eq:eigendecomp}
  \bK = \bV \,\Lambda\, \bV^{-1}, \qquad
  \Lambda = \operatorname{diag}(\lambda_1, \ldots, \lambda_N).
\end{equation}
This yields:

\begin{itemize}
  \item \textbf{Koopman eigenvalues}~$\lambda_j$: encode frequencies
    ($\omega_j = \operatorname{Im}(\ln \lambda_j)$) and growth/decay rates
    ($\sigma_j = \operatorname{Re}(\ln \lambda_j)$).
  \item \textbf{Koopman eigenfunctions:}
    $\varphi_j(\bx) = \bv_j^\top\, \bg_\theta(\bx)$, i.e., linear combinations
    of the snapped EML observables.  Since each $g_i$ is a closed-form elementary
    expression, $\varphi_j$ is also elementary --- fully interpretable.
  \item \textbf{Koopman modes:} $\bm_j = \bC\,\bv_j$, linking each eigenfunction
    to its contribution in state space.
\end{itemize}

\noindent
This is a fundamental advantage over deep-learning Koopman approaches
\citep[e.g., LRAN,][]{otto2019,lusch2018}: instead of opaque neural network
observables, we get symbolic formulas that can be analyzed, simplified, and
compared to physics.


\subsection{GPU Acceleration and Backend Comparison}
\label{sec:cuda}

The EML tree architecture is naturally parallelizable: all $N$ trees at
each level are evaluated in a single batched call, and the Horner-form
Taylor primitives decompose into multiply-add chains.  We benchmark
four computational backends on the Lorenz-63 task (depth~2, $N\!=\!16$,
20{,}000 training pairs, batch size~2048) to measure the realized
speedup from GPU acceleration.

\subsubsection{Backends}

\begin{enumerate}
  \item \textbf{CPU Taylor} --- Horner-form Taylor series on CPU (the
    baseline from all previous experiments).
  \item \textbf{GPU Taylor} --- The same Taylor implementation moved to
    GPU via \texttt{.to("cuda")}.  Each Horner loop step dispatches a
    separate CUDA kernel.
  \item \textbf{GPU Native} --- Replace the Taylor series with
    \texttt{torch.exp}/\texttt{torch.log} (CUDA's built-in math library,
    which uses the same range-reduction and minimax polynomial techniques
    as SLEEF~\citep{sleef}).
  \item \textbf{GPU Compiled} --- Apply \texttt{torch.compile} to fuse
    the exp-minus-log into a single Triton kernel, eliminating intermediate
    memory round-trips.
\end{enumerate}

\subsubsection{Related Libraries}

No native CUDA implementation of the full EML operator exists at the time
of writing.  The closest related work includes:
SLEEF (\url{https://sleef.org/cuda.xhtml}), which provides inlinable CUDA
device functions for $\exp$ and $\ln$ with configurable ULP error bounds
(0.5--3.5~ULP);
PolyKAN \citep{polykan2025}, which implements fused CUDA kernels for
polynomial Kolmogorov--Arnold networks via 2D tiling;
and FlashInfer (\url{https://github.com/flashinfer-ai/flashinfer}), which
contains PTX-level wrappers for \texttt{ex2.approx} and \texttt{lg2.approx}
instructions.  Our GPU Native backend uses CUDA's built-in math functions,
which employ the same mathematical techniques as SLEEF (range reduction
plus minimax polynomials).

\subsubsection{Measured Results}

Table~\ref{tab:timing} reports training throughput, inference latency, and
peak GPU memory for all four backends, measured on an NVIDIA RTX~5070~Ti
(16\,GB, CUDA~12.8) with PyTorch~2.11.

\begin{table}[t]
\centering
\caption{Computational performance comparison (Lorenz-63, depth~2,
  $N\!=\!16$, 720 parameters, 300 training epochs, batch size~2048).
  Training time is mean $\pm$ std over 3 runs.  Inference latency is
  median of 1{,}000 timed forward passes (lift $\to$ predict $\to$
  reconstruct) after 100 warm-up passes.  Rollout time is a 500-step
  iterative prediction.  All GPU measurements include CUDA
  synchronization barriers.}
\label{tab:timing}
\begin{tabular}{lccccc}
\toprule
Backend & \shortstack{Train\\Time (s)} & \shortstack{Train\\(pairs/s)}
  & \shortstack{Infer $B\!=\!1$\\(ms)} & \shortstack{Infer $B\!=\!4096$\\(ms)}
  & \shortstack{Peak\\Mem (MB)} \\
\midrule
CPU Taylor       & $13.5 \pm 0.4$  & 446{,}000   & 0.83 & 34.0  & ---  \\
GPU Taylor       & $3.2 \pm 0.1$   & 1{,}849{,}000 & 2.55 & 2.65 & 37   \\
\textbf{GPU Native}  & $\mathbf{1.8 \pm 0.3}$ & $\mathbf{3{,}379{,}000}$
                     & $\mathbf{0.82}$ & $\mathbf{0.90}$ & 37 \\
GPU Compiled     & $2.4 \pm 1.0$   & 2{,}973{,}000 & 0.88 & 0.96 & 50   \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{GPU Taylor is kernel-launch dominated.}
Moving the Taylor implementation to GPU yields a $4.2\times$ training speedup
($13.5 \to 3.2$\,s), but inference latency is \emph{worse} than CPU at
small batch sizes ($2.55$ vs.\ $0.83$\,ms at $B\!=\!1$) and nearly constant
across batch sizes ($2.55$--$2.65$\,ms).  This is because the Horner-form
evaluation dispatches $d \times (M_e + M_l) \approx 2 \times 28 = 56$
sequential kernel launches per forward pass --- each launch incurs
${\sim}5$--$10\,\mu$s overhead, which dominates at small batch sizes.

\paragraph{GPU Native eliminates Taylor overhead.}
Replacing the Python-level Horner loops with CUDA's built-in \texttt{expf}/\texttt{logf}
reduces the EML operation to two kernel calls per tree level (one for $\exp$,
one for $\ln$), yielding the fastest overall backend:
$7.5\times$ training speedup over CPU ($13.5 \to 1.8$\,s) and
$1.8\times$ over GPU Taylor.  Inference at $B\!=\!4096$ drops from
$2.65$\,ms (GPU Taylor) to $0.90$\,ms --- a $2.9\times$ improvement.

\paragraph{GPU Compiled matches GPU Native after warm-up.}
The \texttt{torch.compile} backend (which fuses exp-minus-log into a single
Triton kernel) achieves nearly identical steady-state performance to GPU
Native: $0.96$\,ms at $B\!=\!4096$ vs.\ $0.90$\,ms.  However, the first
training run incurs a ${\sim}2$\,s compilation overhead ($3.8$\,s vs.\ $1.6$\,s
for subsequent runs), reflected in the higher standard deviation.  Peak
memory is $50$\,MB vs.\ $37$\,MB due to Triton's code cache.

\paragraph{CPU wins at $B\!=\!1$ inference.}
For single-sample inference ($B\!=\!1$), CPU Taylor ($0.83$\,ms) outperforms
GPU Taylor ($2.55$\,ms) due to kernel launch and host--device transfer overhead.
GPU Native ($0.82$\,ms) matches CPU because the reduced kernel count amortizes
the launch cost.  This has practical implications for real-time control
applications where inference is one-sample-at-a-time.


\subsubsection{Batch Size Scaling}

Table~\ref{tab:batch-scaling} shows inference latency across batch sizes
for all four backends.

\begin{table}[t]
\centering
\caption{Inference latency (ms) vs.\ batch size.  CPU scales linearly;
  GPU Taylor is flat (kernel-launch dominated); GPU Native and Compiled
  are nearly flat with marginal increase at large batches.  Values are
  medians of 1{,}000 timed passes after 100 warm-up passes.}
\label{tab:batch-scaling}
\begin{tabular}{lrrrrr}
\toprule
Backend & $B\!=\!1$ & $B\!=\!64$ & $B\!=\!256$ & $B\!=\!1024$ & $B\!=\!4096$ \\
\midrule
CPU Taylor      & 0.83 & 1.51 & 3.28 &  9.79 & 33.97 \\
GPU Taylor      & 2.55 & 2.63 & 2.63 &  2.64 &  2.65 \\
GPU Native      & \textbf{0.82} & \textbf{0.90} & \textbf{0.90} & \textbf{0.90} & \textbf{0.90} \\
GPU Compiled    & 0.88 & 0.97 & 0.97 &  0.98 &  0.96 \\
\bottomrule
\end{tabular}
\end{table}

\noindent
The CPU backend's linear scaling ($0.83 \to 34$\,ms from $B\!=\!1$ to
$B\!=\!4096$) reflects sequential processing of batch elements.  GPU
backends process all batch elements in parallel, so latency is roughly
constant until the GPU's streaming multiprocessors saturate.  The
crossover where GPU Native overtakes CPU occurs between $B\!=\!1$ and
$B\!=\!64$, making GPU the preferred backend for any practical batch
size in online prediction or multi-trajectory training.

\subsubsection{Implications for CUDA Kernel Development}

The $2.9\times$ inference speedup from GPU Taylor to GPU Native
demonstrates that the Horner-form Taylor loops are the primary bottleneck
on GPU, not the arithmetic itself.  A hand-written fused CUDA kernel
(as described in the existing SLEEF and PolyKAN literature) would need
to beat GPU Native's $0.90$\,ms at $B\!=\!4096$, which is already
close to the hardware floor for 16 dictionary evaluations over a
depth-2 tree.  The marginal benefit of custom kernels is therefore
limited for shallow trees but may become significant at depth~$\ge 4$
where the number of sequential kernel launches grows exponentially.


\subsection{Limitations and Future Work}
\label{sec:limitations}

The progressive depth analysis (Section~\ref{sec:depth-analysis}) revealed
that naively increasing tree depth degrades both E1 and E2, contradicting the
expectation that a richer function space should improve approximation quality.
This identifies several concrete limitations and motivates targeted
experiments.

\paragraph{Limitation 1: Gradient attenuation in deep EML trees.}
The formula collapse at depth~$\ge 3$ (Table~\ref{tab:depth-formulas})
indicates that gradient signal fails to propagate through nested clamped
exponentials.  Each EML node applies $\exp(\text{clamp}(x))$, whose gradient is
$\exp(\text{clamp}(x))$ when $|x| < C$ and zero otherwise.  At depth~$d$,
the backward pass multiplies $d$ such terms, creating either exponential
gradient magnitudes (when activations are large) or vanishing gradients (when
the clamp saturates).  The tight clamp required for numerical stability
($C\!=\!3$--$5$) exacerbates this by restricting the active gradient region.

\paragraph{Limitation 2: Structural E2 gap.}
The 65-point E2 gap to EDMD-Poly is not a training failure but a fundamental
expressiveness mismatch: depth-2 EML trees cannot represent polynomials, and the
exponential/logarithmic observables they produce have the wrong spectral
signature for long-term Lorenz statistics.

\paragraph{Limitation 3: Initialization sensitivity.}
All runs used random initialization.  The depth-2 success suggests that shallow
trees benefit from a relatively smooth loss landscape where Gumbel-softmax
exploration suffices.  Deeper trees likely have many more local minima,
making random initialization inadequate.

\subsubsection{Suggested Further Experiments}

The following experiments target the specific bottlenecks identified above:

\begin{enumerate}
  \item \textbf{Curriculum depth growth.}  Train at depth~2 until convergence
    (Phase~3 snap), then grow each tree by appending new leaf nodes with random
    logits and resume training with the routing structure from the shallower
    tree frozen.  This provides a strong initialization for the deeper tree
    and avoids the cold-start problem that plagues depth-3+ random
    initialization.  The experiment should sweep growth schedules:
    $2 \to 3$, $2 \to 3 \to 4$, and $2 \to 4$ (skipping depth~3).

  \item \textbf{Hybrid EML+Polynomial dictionary.}  Augment the EML observable
    dictionary with polynomial features (monomials up to degree~4) to directly
    address the E2 gap.  The combined dictionary
    $\bg(\bx) = [\text{EML}_1(\bx), \ldots, \text{EML}_N(\bx),\;
    x_0^2, x_0 x_1, \ldots]$ would let the Koopman matrix leverage polynomial
    features for long-term spectral fidelity while retaining EML's symbolic
    interpretability for the learned components.  This is straightforward to
    implement within the PySINDy \texttt{ConcatLibrary} framework.

  \item \textbf{Gradient-aware clamping.}  Replace hard clamping with a smooth
    saturation function (e.g., $C \cdot \tanh(x/C)$) that preserves nonzero
    gradients everywhere.  This should reduce gradient attenuation at depth~3+
    while still bounding $\exp$ inputs.  Alternatively, apply gradient clipping
    \emph{per tree level} rather than globally, allowing the optimizer to
    maintain signal through deeper layers.

  \item \textbf{EDMD bootstrap initialization.}  Run EDMD-Poly first, identify
    the top-$k$ polynomial observables by Koopman reconstruction error, then
    initialize EML tree routing to approximate those polynomials (using known
    EML representations of monomials).  This warm-starts the optimizer near a
    good solution, which may be especially beneficial at depth~$\ge 5$ where
    polynomial representations become expressible.

  \item \textbf{Spectral loss augmentation.}  Add an explicit spectral matching
    term to the training loss:
    $\cL_{\text{spec}} = \|\log P_{\text{pred}}(f) - \log P_{\text{truth}}(f)\|^2$
    over the first $k$ frequency bins.  This directly optimizes for E2 during
    training rather than relying on the one-step prediction loss to implicitly
    capture long-term statistics.

  \item \textbf{Complex-valued EML for oscillatory systems.}  As demonstrated
    in Section~\ref{sec:complex-eml}, a mixed real/complex dictionary recovers
    most short-term accuracy while adding oscillatory observables.
    Further work could explore adaptive real/complex tree ratios,
    complex-aware spectral losses, and applications to systems with
    dominant oscillatory modes (e.g., coupled oscillators, fluid flows).

  \item \textbf{EML-KAN hybrid architectures.}  Replace KAN's B-spline
    activation functions \citep{liu2025kan} with EML nodes, combining KAN's
    universal approximation structure with EML's symbolic snap.  The
    Kolmogorov--Arnold representation theorem guarantees that any continuous
    function can be decomposed into univariate functions composed with
    addition; if each univariate function is an EML tree, the entire network
    snaps to a symbolic expression.
\end{enumerate}


% ============================================================================
\subsection{Complex EML Primitives}
\label{sec:complex-eml}

The Lorenz system possesses complex Koopman eigenvalues corresponding to
oscillatory dynamics on the attractor.  To test whether the EML grammar
benefits from access to trigonometric functions, we extend the routing
candidate set with imaginary-valued primitives and compare three
configurations:

\begin{enumerate}
  \item \textbf{Real baseline} --- candidates $\{1, x_0, x_1, x_2\}$ (original grammar).
  \item \textbf{i-only} --- candidates $\{1, i, x_0, x_1, x_2\}$; adds the
    imaginary unit as a selectable leaf.
  \item \textbf{i\,+\,ix} --- candidates $\{1, i, x_0, x_1, x_2, ix_0, ix_1, ix_2\}$;
    adds both the imaginary unit and imaginary-scaled state variables.
\end{enumerate}

\noindent
In complex mode the Koopman matrix~$K$ is complex-valued, and
reconstruction uses $\hat{x} = C_{\text{re}}\,\Re(g) + C_{\text{im}}\,\Im(g)$
to map complex observables back to real state space.
The prediction loss becomes
$\cL_{\text{pred}} = \tfrac{1}{N}\sum_i |g_{k+1,i}^{\text{pred}}
  - g_{k+1,i}^{\text{true}}|^2$,
where $|\cdot|$ is the complex modulus.

\begin{table}[t]
\centering
\caption{Complex EML comparison on Lorenz-63 CTF metrics (depth~2,
  $N\!=\!16$). Initial experiments (top) show depth collapse; four
  depth-collapse fixes are evaluated below. D2 = number of trees
  retaining depth-2 composition after snapping.}
\label{tab:complex-eml}
\begin{tabular}{lrrrrrr}
\toprule
Config & E1 & E2 & RMSE & Steps & D2 & Params \\
\midrule
\multicolumn{7}{l}{\emph{Initial experiments (v1)}} \\
Real baseline           &  \textbf{7.92} & $-21.71$ & \textbf{0.886} & \textbf{142} & 16 &   720 \\
i\,+\,ix original      & $-1.52$ & $-18.32$ & 0.977 & 111 &  0 & 1{,}152 \\
\midrule
\multicolumn{7}{l}{\emph{With depth-collapse fixes (v2)}} \\
+ child-logit bias      & $-0.00$ & $-215$ & 0.962 & 126 & 16 & 1{,}152 \\
+ warm-start            & $-0.05$ & $-57.8$ & 0.963 & 126 & 16 & 1{,}152 \\
\textbf{+ mixed dict}   &   1.89  & $-22.9$ & 0.944 & 135 & 11 &   960 \\
+ slow anneal           & $-2.53$ & $-58.1$ & 0.987 & 107 &  0 & 1{,}152 \\
\textbf{+ all combined}  &   2.22  & $\mathbf{-18.9}$ & 0.941 & 137 & 16 & 960 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Depth collapse under complex grammar expansion.}
Initial experiments revealed that complex-mode trees collapsed to depth-1
formulas (D2\,$= 0/16$), losing the nested composition that powered the
real baseline.  The root cause is \emph{candidate dilution}: with 8 leaf
candidates (vs.\ 4 for real), the Gumbel-softmax probability mass for the
child-node slot is halved during soft exploration, causing the optimizer
to settle into depth-1 local minima before annealing can harden the child
selection.

\paragraph{Four depth-collapse fixes.}
We tested: (1)~a child-logit initialization bias of $+2.0$ on non-leaf
levels; (2)~warm-start training (800 real epochs, then expand logits);
(3)~a mixed dictionary with half real, half complex trees; and
(4)~extended exploration phase ($75\%$ of training at high~$\tau$).
Table~\ref{tab:complex-eml} shows that the child-logit bias~(1) and
warm-start~(2) both recover depth-2 composition (D2\,$=16$) but at the
cost of catastrophically worse E2, because they force premature
composition of poorly-routed complex leaf nodes.  Slow annealing~(4)
alone does not prevent collapse.

\paragraph{Mixed dictionary is the recommended approach.}
The mixed dictionary~(3) uniquely preserves short-term accuracy
(E1\,$= 1.89$, 135 valid steps) while adding complex observables.
The real half maintains depth-2 composition with the same
$\eml(\eml(1,x_i), x_j)$ patterns found in the pure-real baseline;
the complex half discovers oscillatory forms such as
$\eml(i, ix_1)$ and $\eml(i, \eml(i, x_0))$.
The combined configuration (all fixes, E1\,$= 2.22$, E2\,$= -18.9$)
achieves the best balance among depth-2 complex models.

\paragraph{Depth-2 complex trees do not improve E2 over depth-1.}
A surprising finding is that the v1 depth-1 complex formulas
($\text{E2} = -18.3$) outperform all v2 depth-2 complex configurations.
Nested complex exponentials such as $\exp(\exp(ix))$ produce
ill-conditioned spectral signatures, whereas shallow
$\exp(ix_j) - \ln(x_k)$ captures the attractor's oscillatory structure
more faithfully.


% ============================================================================
\section{Conclusion}
\label{sec:conclusion}

We have presented a framework for learning Koopman operator representations using
EML observable dictionaries --- parameterized binary trees of the universal
$\eml(x,y) = \exp(x) - \ln(y)$ operator.  The completeness of EML over elementary
functions provides a principled, gradient-trainable search space that is guaranteed
to contain Koopman eigenfunctions whenever they are elementary.  GPU-optimized
Horner-form Taylor primitives eliminate transcendental function calls, and
Gumbel-softmax annealing with three-phase training produces exact symbolic
observables upon convergence.

A complete evaluation across all 12 CTF metrics on the Lorenz-63 system yielded
an average score of $-3.71$, placing EML-Koopman within one point of SINDy
($-3.00$) and 16 points above PyKoopman ($-20.11$) --- competitive with the best
interpretable methods while using only 720 parameters.  Beyond forecasting, the
full CTF evaluation revealed an unexpected strength: the Koopman
lift$\to$reconstruct pipeline provides effective denoising (E3 $= 33.68$,
E5 $= 30.65$), and the compact model size confers natural robustness under data
scarcity (E7 $= 7.77$ with 10\% training data).  Every discovered observable is a
closed-form symbolic expression, yielding interpretable Koopman eigenfunctions that
black-box methods cannot produce.

A systematic depth sweep across 11 configurations (depths 2--5, 16--32
observables) revealed that deeper trees do not improve performance from random
initialization: E1 degrades from 7.70 (depth~2) to as low as 1.86 (depth~3),
and E2 remains negative at all depths.  Inspection of snapped formulas shows
that depth-3+ trees collapse to root-only expressions, indicating gradient
attenuation through nested clamped exponentials.  The persistent negative
long-term spectral scores (E2, E4, E6, E8) across all conditions reflect a
structural mismatch: depth-2 EML trees cannot express polynomials, whose spectral
signatures dominate Lorenz long-term statistics.  Parametric generalization
(E11, E12) remains weak without explicit parameter conditioning.

Extending the grammar with complex-valued primitives (Section~\ref{sec:complex-eml})
revealed a depth-collapse phenomenon: the expanded candidate set dilutes
Gumbel-softmax routing, causing trees to flatten to depth~1.  A mixed
dictionary (half real, half complex trees) recovers E1 to 1.89 while
gaining access to oscillatory observables, and the combined-fixes
configuration achieves the best complex E2 ($-18.9$) with full depth-2
composition.

These findings point to four high-priority directions: (1)~curriculum depth
growth, where shallow-trained trees provide warm-start initialization for
deeper structures; (2)~hybrid EML+polynomial dictionaries that combine
symbolic interpretability with spectral fidelity; (3)~gradient-aware
clamping via smooth saturation to maintain signal through deep trees;
and (4)~parameter-conditioned Koopman matrices $\bK(\rho)$ to enable
generalization across bifurcation parameters.
A systematic computational benchmark (Section~\ref{sec:cuda}) on an RTX~5070~Ti
demonstrated $7.5\times$ training speedup from CPU to GPU with native math
functions, and identified kernel launch overhead from the Python-level Horner
loops as the primary GPU bottleneck --- resolved by switching to CUDA's built-in
\texttt{exp}/\texttt{log} or \texttt{torch.compile} fusion.

Joint real-complex training that combines real-grammar short-term accuracy with
complex-grammar spectral fidelity, EML-KAN hybrid architectures, and deeper
kernel fusion for depth-$\ge 4$ trees remain promising longer-term directions.


\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
